Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an Empirical Study of Strategies in the Moran process for the Iterated Prisoner's Dilemma

### 2018-07-18 main.tex
Fix unclear sentences.
Also took the opportunity for minor fixes.

The notion of a handshake has been described previously in~\cite{Robson1990} and
corresponds to idea that an individual exhibits behaviour that starts with a
recognisable pattern. This in turn allows them to identify memers of their own
type (who would exhibit the same pattern). This is analogous
to the biological notion of `kin recognition' where individuals have the ability
to recognise the phenotype of their own kin ~\cite{Mateo2003}.

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider relatively simple strategies. A few works
looked at evolutionary stability of agent-based strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies: strategies that recall
the events of the previous round only.

Of particular interest are the zero determinant strategies introduced in
\cite{Press2012}. It was argued in \cite{stewart2013extortion} that generous
ZD strategies are robust against invading strategies. However, in \cite{Lee2015},
a strategy using machine learning techniques was capable of resisting invasion
and also able to invade any memory one strategy. In 2017, \cite{Hilbe2017}
has investigated the effect of memory length on strategy performance and the
emergence of cooperation but this is not done in a Moran process context and only
considers specific cases of memory 2 strategies. In \cite{Adami2013} it was
recognised that many zero determinant strategies do not fare well against
themselves. This is a disadvantage for the Moran process where the best
strategies cooperate well with other players using the same strategy.

@@

To carry out this numerical experiment, 164
strategies, listed (with their properties) in the Appendix,
are used from the Axelrod library. The appendix includes citations to the
original description of each strategy. There are
43 stochastic and
121 deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponent's actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

@@

A number of these strategies have been trained this way (see~\cite{Harper2017})
prior to this study and not specifically for the Moran process.

@@

One particular family of strategies that has been studied in the literature are
called finite state machines. These mathematical models consist of states and
and responses to actions from one state to another. In the context of the IPD, a
finite state machine, is a mapping from states and opponent actions to states
and an action. For further details on this type of strategy, the reader is
referred to~\cite{Ashlock2006, Harper2017}.

There are three further strategies trained specifically for this
study; Trained FSM 1, 2, and 3 (TF1 - TF3). These are finite state
machines of 16, 16, and 8 states respectively.  These are shown in
Fig~\ref{fig:tf1},~\ref{fig:tf2} and~\ref{fig:tf3}, using the notation common in
the literature where \(A_1/A_2\) is the action of the opponent \(A_1\) and the
response of the player \(A_2\) as well as arrows corresponding to changes of
state.

@@

As opposed to the previously described strategies, these strategies were trained
with the objective function of \textit{mean fixation probabilities for Moran
processes} starting at initial population states consisting of \(N/2\)
individuals of the training candidates and \(N/2\) individuals of an opponent
strategy, taken from a selection of 150 opponents from the Axelrod library:

@@

The next section will further validate the methodology by comparing
simulated results to analytical results in a few selected cases. The main
results of this
manuscript are presented in which will
present a detailed analysis of all the data generated. Finally,
a discussion and conclusion will offer future avenues for the work
presented here.

@@

It is also noted that TF1, TF2 and TF3 all perform well. This is also the \(N\)
for which a zero determinant strategy does appear in the top 10\% ranking
strategies: ZD-extort-4. The performance of zero determinant strategies will be
examined more closely.

@@

% In the next sections close attention to
% strategies who are strong invaders/resistors is given.

@@

There are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF1 and TF2. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshakes of CS and
Handshake (which ranks highly for the smaller values of \(N\)) were
programmed, the handshakes of TF1 and TF2 evolved through an evolutionary
process without any priming.

@@

The results for \(x_{N/2}\) show similarities to the results for \(x_{N-1}\) and
in particular TF1, TF2 and TF3 ranked one, three and eight. This is to be
expected since, as described previously these strategies
were trained in an initial population of \((N/2, N/2)\) individuals.

@@

Fig~\ref{fig:correlation_coefficients} shows the correlation coefficients
of the ranks of strategies in differing population size. How well a strategy
performs in any Moran process for \(N>2\) has
little to do with the performance for \(N=2\). This illustrates why the strong
performance of zero determinant strategies predicted in \cite{Press2012} does
not extend to larger populations. This was discussed theoretically in
\cite{Adami2013} and observed empirically in these simulations.

>>>>

The notion of a \textbf{handshake} has been described previously
in~\cite{Robson1990} and corresponds to the idea that an individual exhibits
behaviour that starts with a recognisable pattern. This in turn allows them to
identify individuals of their own type (who would exhibit the same pattern).
This is analogous to the biological notion of `kin recognition' where
individuals have the ability to recognise the phenotype of their own
kin~\cite{Mateo2003}.

@@

The Moran process was initially introduced in~\cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer~\cite{West2016} and the
emergence of cooperative behaviour in spatial topologies~\cite{Nowak2017}.
However these works mainly consider relatively simple strategies. A few works
looked at evolutionary stability of agent-based strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In~\cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies: strategies that recall
the events of the previous round only.

Of particular interest are the zero determinant strategies introduced in
\cite{Press2012}. It was argued in~\cite{stewart2013extortion} that generous
ZD strategies are robust against invading strategies. However, in~\cite{Lee2015},
a strategy using machine learning techniques was capable of resisting invasion
and also able to invade any memory one strategy. In 2017,~\cite{Hilbe2017}
has investigated the effect of memory length on strategy performance and the
emergence of cooperation but this is not done in a Moran process context and only
considers specific cases of memory 2 strategies. In~\cite{Adami2013} it was
recognised that many zero determinant strategies do not fare well against
themselves. This is a disadvantage for the Moran process where the best
strategies cooperate well with other players using the same strategy.

@@

To carry out this numerical experiment, 164
strategies, listed (with their properties) in the Appendix,
are used from the Axelrod library. The appendix also includes citations to the
original description of each strategy. There are
43 stochastic and
121 deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponent's actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

@@

A number of these strategies have been trained this way (see~\cite{Harper2017})
prior to this study and not specifically for the Moran process. For example:

@@

One particular family of strategies that has been studied in the literature are
called finite state machines. These mathematical models consist of states and
and responses to actions which indicate a next state given an action.  In the
context of the IPD, a finite state machine, is a mapping from an arbitrary list
of states and opponent actions (Cooperation or Defection) to states and an
action (Cooperation or Defection). For further details,
the reader is referred to~\cite{Ashlock2006, Harper2017}.

There are three further strategies trained specifically for this
study; Trained FSM 1, 2, and 3 (TF1 \- TF3). These are finite state
machines of 16, 16, and 8 states respectively.  These are shown in
Fig~\ref{fig:tf1},~\ref{fig:tf2} and~\ref{fig:tf3}, using the notation common in
the literature where \(A_1/A_2\) is the action of the opponent \(A_1\) and the
response of the player \(A_2\) as well as arrows corresponding to changes of
state.

@@

As opposed to the previously described strategies~\cite{Harper2017}, these
strategies were trained with the objective function of \textit{mean fixation
probabilities for Moran processes} starting at initial population states
consisting of \(N/2\) individuals of the training candidates and \(N/2\)
individuals of an opponent strategy, taken from a selection of 150 opponents
from the Axelrod library:

@@

The next section will further validate the methodology by comparing
simulated results to analytical results in a few selected cases. The main
results of this
manuscript will
present a detailed analysis of all the data generated. Finally,
a discussion and conclusion will offer future avenues for the work
presented here.

@@

It is also noted that TF1, TF2 and TF3 all perform well for this case of
\(N=2\). This is also the value of \(N\) for which a zero determinant strategy
does appear in the top 10\% ranking strategies: ZD-extort-4. The performance of
zero determinant strategies will be examined more closely.

@@

@@

There are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF1 and TF2. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshakes of CS and
Handshake (which ranks highly for the smaller values of \(N\)) were
programmed, the handshakes of TF1 and TF2 evolved without any priming.

@@

The results for \(x_{N/2}\) show similarities to the results for \(x_{N-1}\) and
in particular TF1, TF2 and TF3 ranked one, third and eighth. This is to be
expected since, as described previously these strategies
were trained in an initial population of \((N/2, N/2)\) individuals.

@@

Fig~\ref{fig:correlation_coefficients} shows the correlation coefficients
of the ranks of strategies in differing population size. How well a strategy
performs in any Moran process for \(N>2\) has
low correlation with the performance for \(N=2\). This illustrates why the strong
performance of zero determinant strategies predicted in~\cite{Press2012} does
not extend to larger populations. This was discussed theoretically in
\cite{Adami2013} and observed empirically in these simulations.
